Signatures of Foreground Sky in Power Spectra of Redshifted Neutral Hydrogen from the Epoch of Reionization

### 2014-08-29 foreground_delay_spectrum_modeling.tex
From v0.1: =========
(*) Edited first half of the foreground_power_spectra_modeling.tex file upto and
including section 6.
(*) Removed unnecessary .eps files from the version and use only those needed to
compile.

We characterize signatures of foreground radio emission and instrument configuration on the observed power spectra of redshifted H{\sc i} 21~cm line emission from the epoch of reionization (EoR) using a baseline (antenna pair) based ``delay spectrum'' analysis technique. We use the 128-tile Murchison Widefield Array (MWA) instrument configuration for our studies. Our analysis is centered around 185~MHz with 30.72~MHz bandwidth at 40~kHz frequency resolution. We simulate the delay spectrum response of interferometers with antenna pair spacings up to $\sim 315$~m, for the first time, to an all-sky foreground model consisting of diffuse (on scales $\gtrsim\,$0\fdg 85) and compact Galactic and extragalactic emission. By matching simulated parameters to those of a subset of first season of observing with the MWA, we identify and verify key foreground and instrumental signatures seen in data. A wedge-shaped region of foreground contamination in the delay spectra is confirmed. There is a spillover from this contamination into relatively foreground--free regions ({\it EoR window}) in the delay spectrum caused due to the spectral properties of foreground emission and the instrument. Due to projection effects, diffuse emission produces edge-heavy features. Features in the inner regions of the wedge are predominantly due to compact foreground objects. This results in a ``pitchfork'' shaped delay spectrum. We provide a practical diagnostic tool for scheduling EoR observations that helps in reducing foreground contamination.

@@

After the CMB epoch, the period in the history of the universe often referred to as the {\it dark ages} was followed by the epoch of reionization (EoR). This was a period of non-linear growth of density perturbations and astrophysical evolution. Studying the epoch of reionization holds the key to understanding this evolution. Recently, oberving the redshifted 21~cm spin transition of neutral hydrogen has emerged as one of the most promising experiments to fill the gaps in our understanding of the universe's history.

@@

In this paper, in one of the most detailed characterizations to date, using an interferometer (antenna pair) based delay spectrum technique, we demonstrate the signatures of unsubtracted diffuse and compact foreground models in delay spectrum, a quantity closely related to the sought power spectrum. We explore the diversity in these signatures as a function of observation and instrument parameters and confirm our findings with data obtained from the MWA.

We have organized the rest of the paper as follows. \S\ref{sec:delay-spectrum} defines the delay spectrum and its relation to the power spectrum. \S\ref{sec:instrument} and \S\ref{sec:obsparms} describe the MWA instrumental parameters and observation parameters used in this study respectively. In \S\ref{sec:data-analysis}, we describe our data analysis. In \S\ref{sec:modeling}, we describe the building blocks of our delay spectrum modeling including the foreground models used. Here, we also establish that the results of modeling and data analysis match with each other. In \S\ref{sec:delay-spectrum-analysis}, we present a detailed analysis of the signatures of foregrounds on the modeled delay spectrum and its dependence on instrumental and observation parameters. In \S\ref{sec:fg-grading}, we use our understanding to grade the delay spectrum as a function of local sidereal time (LST) in order to obtain a useful diagnostic for planning future observations. In \S\ref{sec:summary}, we present our summary and conclusions.

@@

Interferometer array data known as {\it visibilities}, $V_f(\overline{\mathbf{u}},f)$, are correlations between electric fields measured at different frequencies ($f$) by different antenna pairs with separation vectors $\overline{\mathbf{x}}$, where $\overline{\mathbf{u}}\equiv \overline{\mathbf{x}}f/c$ and $c$ is the speed of light. $\overline{\mathbf{u}}$ is related to transverse spatial frequency modes ($\overline{\mathbf{k}}_\perp$) of brightness distribution of the sky as $\overline{\mathbf{k}}_\perp \equiv 2\pi\overline{\mathbf{u}}/D(z)$, where $D(z)$ is the transverse comoving distance at a redshift $z$. If $I(\overline{\mathbf{l}},f)$ is the emission at different frequencies on the sky as a function of direction unit vector ($\overline{\mathbf{l}}$) specified as direction cosines, $V_f(\overline{\mathbf{u}},f)$ represents the Fourier decomposition of $I(\overline{\mathbf{l}},f)$ attenuated by the antenna's angular power pattern $A(\overline{\mathbf{l}},f)$, into transverse spatial frequency components. Taking into account the instrumental bandpass weights $W_f(f)$, this relation can be written as:
\begin{align}\label{eqn:obsvis}
  V_f(\overline{\mathbf{u}},f) &= \int A(\overline{\mathbf{l}},f)\,I(\overline{\mathbf{l}},f)\,W_f(f)\,e^{-i2\pi \overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}\,\dif^2 \overline{\mathbf{l}}.
\end{align}

We define the {\it delay spectrum} $V_\eta(\overline{\mathbf{u}},\eta)$ to be the inverse Fourier transform of $V_f(\overline{\mathbf{u}},f)$ along the frequency coordinate:
\begin{align}\label{eqn:delay-transform}
  V_\eta(\overline{\mathbf{u}},\eta) &= \int V_f(\overline{\mathbf{u}},f)\,W'_f(f)\,e^{i2\pi\eta f}\,\dif f,
\end{align}
where, $W'_f(f)$ is a frequency window weighting function which can be chosen to control the quality of the delay spectrum \citep{thy13,ved12}. $\eta$ represents the signal delay between the antenna pairs in an interferometer given by:
\begin{equation}\label{eqn:delay}
  \eta = \frac{\overline{\mathbf{x}}\cdot\overline{\mathbf{l}}}{c} = \frac{\overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}{f}.
\end{equation}
We use the terms delay and lag interchangeably in the rest of this paper.

Since we are studying a redshifted H{\sc i} spectral line, $f$ is a measure of cosmological distance along the line of sight. Consequently, $\eta$ is a measure of line of sight spatial frequency, $k_\parallel$. Thus, $|V_\eta(\overline{\mathbf{u}},\eta)|^2$ is directly related to the spatial power spectrum of redshifted H{\sc i} distribution, $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$. However, foreground objects emit in radio frequencies and contaminate the signals from the redshifted H{\sc i}. Due to their spatial and spectral properties, $V_\eta(\overline{\mathbf{u}},\eta)$ also contains contamination from foreground radio sources \citep{thy13,tro12,mor12,bow09}. These contaminations have to be characterized precisely in order to reduce their impact on redshifted H{\sc i} power spectrum detection sensitivity. We, heareafter, examine in detail the delay spectrum of emission from unsubtracted foreground objects and compare with preliminary observations with the MWA.

@@

We use the 128-tile layout of the MWA \citep{bea12}. We choose a center frequency of 185~MHz with a bandwidth of 30.72~MHz divided into 768 frequency channels each of width 40~kHz. The choice of center frequency translates to $z\approx 6.68$ for the 21~cm spin flip transition of H{\sc i}. The bandwidth was chosen to be close to the maximum instantaneous value of the MWA in order to maximize the resolution in delay coordinates and obtain detailed characteristics of delay spectra from foregrounds. We restrict our analysis to the innermost 2000 pairs covering $|\overline{\mathbf{x}}| \leq x_\textrm{max} \approx 315$~m, or $|\overline{\mathbf{u}}| \leq u_\textrm{max} \approx 157.5$ at a frequency of 185~MHz. These interferometers are responsive to transverse spatial structures on scales $\gtrsim 1/u_\textrm{max} \approx 22$\arcmin.

The bandpass weights in the frequency window of the MWA are obtained through a 8-tap Polyphase Filter Bank (PFB) using a Kaiser window with the $\beta$ parameter set to 5. The bandpass weights consist of coarse frequency channels of width 1.28~MHz each of which contains 32 finer frequency channels of width 40~kHz. Our analysis, therefore, contains 24 coarse channels. As the data are subsequently bandpass calibrated, the bandpass weights are normalized to unity. But in the edges of the coarse channels where the bandpass correction factors are $\gtrsim 3$, this correction makes the edge channels signficantly noisy compared to the rest of the spectrum. Thus, we remove 2 fine channels from each side of the coarse channel edges from our data. This constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}. The bandpass properties used in this paper are illustrated in figure~\ref{fig:bandpass}.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/bandpass_properties}
\caption{Bandpass properties used in modeling and analysis of delay spectrum of MWA visibilities. The filled circles joined by dotted lines show the 8-tap PFB bandpass shapes obtained using a Kaiser window with parameter $\beta=5$. The {\it plus} symbols joined by dashed lines are the bandpass correction factors. The dot-dashed line represents the threshold for bandpass correction factor above which frequency channels are flagged. The solid line shows the flagged and gain--corrected bandpass shape, $W_f(f)$. These properties hold for the entire bandwidth of 30.72~MHz centered around 185~MHz. \label{fig:bandpass}}
\end{figure}

Thermal noise in simulated visibilities is estimated assuming a system temperature of $T_\textrm{sys}=440$~K \citep{bow06} for all frequency channels for all interferometers throughout the course of observation. We take into account the increase in thermal uncertainty in frequency channels that results due to aforementioned bandpass correction.

The power pattern of a MWA tile, $A(\overline{\mathbf{l}},f)$ in equation~\ref{eqn:obsvis}, over the entire hemisphere is modeled based on a 4$\times$4 phased array of half-wave dipoles. Hereafter, we use the terms tile and antenna interchangeably to refer to the MWA tile.

@@

One of the primary targets for EoR observations with the MWA is a patch of sky centered around RA = 00h 00m 00.00s, Dec = -30\arcdeg$\,$00\arcmin$\,$00\farcs 0. This is one of the regions of low foreground emission within 27\arcdeg$\,$ from the zenith. The MWA tracks a patch of sky through beams formed and steered electronically by changing delay settings between a phased array of dipoles in a MWA tile. These beamformer delay settings can be changed only in discrete steps. Accordingly, observations were made in a ``drift and shift'' (D-S) mode. Figure~\ref{fig:pointings} shows the direction of pointing of MWA tiles during the course of the observation. The sky is allowed to drift for a certain period of time (usually $\sim 30$~mins) before the discrete shifts in the beamformer delay settings position the beam to be centered again on the patch of sky. This process is repeated throughout the course of the observation $\approx 4.86$~hours.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/dns_pointings}
\caption{MWA tile pointing directions during the course of the observation. The x-axis refers to the Local Sidereal Time (LST) in hours. The axis on the left refers to longitudes, namely, Right Ascension (RA) and Hour Angle (HA) in degrees. Negative values of RA, HA, and LST are to be interpreted as having been wrapped around by 360\arcdeg or 24 hours. The axis on the right refers to the declination of the pointing direction. The RA, HA, and declination are plotted with solid, dashed and dotted line styles respectively. A ``drift and shift'' scheme is used. The sky is allowed to drift for $\sim 30$~mins before the beamformer delay settings change in a discrete step to center the beam around RA = 00h 00m 00.00s, Dec = -30\arcdeg$\,$00\arcmin$\,$00\farcs 0. \label{fig:pointings}}
\end{figure}

@@

{\bf Describe the data analysis here.}

Throughout this paper, $\overline{\mathbf{u}}$ is assumed to be on a coordinate system aligned with the local east and local north directions at the MWA site. For the conventions of Fourier transform and its inverse we use in this paper, in accordance with equation~\ref{eqn:delay}, signals from the sky towards east and north are recorded with positive delays on eastward and northward baselines respectively. For instance, at the beginning of the observation, the Galactic center is on the far west just about to set. Signals from this direction are recorded with a negative delay on interferometers oriented towards the east. Similarly, signals from directions eastward of the local meridian arrive with positive delays on eastward interferometers. In subsequent discussions, unless mentioned otherwise, all the snapshot visibilities are assumed to be phased to zenith.

For purposes of geometrical intuition, we restrict the orientation ($\theta_b$ measured anti-clockwise from East) of all baselines to lie in the range $-67\fdg 5 \leq \theta_b < 112\fdg 5$. Baselines in the other half-plane measure conjugate visibilities with delays of equal magnitude but of opposite sign and hence are ignored in our analysis.

@@

Subsequently, we obtain the delay spectrum of these visibilities using equation~\ref{eqn:delay-transform} while choosing $W'_f(f)$ to be a {\it Blackman--Nuttall} window function for its superior sidelobe suppression as described in \citet{thy13}. Due to periodic gaps in frequency bandpass occurring in intervals of 1.28~MHz (see figure~\ref{fig:bandpass}), the delay spectrum is expected to contain harmonics of actual foreground emission repeated at intervals of 0.78~$\mu$s. Hence, we have employed deconvolution by {\it CLEAN} algorithm \citep{tay99} along the delay axis \citep{par09,par12} to rid the delay spectra of such artifacts. The convolving kernel for the algorithm is given by the inverse Fourier transform of the instrumental bandpass shape.

\subsection{Delay Spectrum from Data}\label{sec:data-delay-spectrum}

Put up plots here and discuss them.

Despite the deconvolution, we still see leakage beyond the cleaned regions indicating the inability of the deconvolution algorithm to fit the data perfectly. This is predominant at short baselines. This is because the maximum delay envelope (boundary of the wedge) consists of mixed emission from large portions of the sky into a narrow range of delays and the deconvolution algorithm does not have sufficient support in delay space to fit the delay spectrum. However, we note that the purpose of deconvolution is to rid the delay spectra of instrumental artifacts to the best extent possible in order to see the underlying signatures of foreground emission. It is not our intention in this paper to use the deconvolution as a foreground removal tool. Hence, our discussion and conclusions still hold despite imperfections in deconvolution.

@@

Below we describe the foreground model we have used in modeling the measured delay spectra.

@@

An instrument such as the MWA has a very wide field of view ($\sim 30$\arcdeg at 185~MHz) for imaging purposes. In the context of EoR H{\sc i} power spectrum it has been shown that unsubtracted forground sources anywhere in the horizon directly contaminate the spatial frequency modes in the power spectrum. In addition, this contamination also spills over into the relatively cleaner regions, called {\it EoR window}, due to spectral properties of the instrument and the foregrounds \citep{thy13,pob13,ved12,par12}. Thus, it is important to consider an all-sky model for foreground objects in evaluating the features seen in the power spectrum.

\citet{bea13} estimated the sensitivity of the MWA to EoR H{\sc i} power spectrum detection taking into account theremal noise effects in the {\it Eor window}. \citet{thy13} estimated the sensitivity taking into account the spillover from foreground contamination from unsubtracted extragalactic point sources in the {\t EoR window} besides thermal noise. In this paper, we include the diffuse Galactic emission for a comprehensive understanding of foreground signatures in the power spectrum. We use an all-sky foreground emission model that consists of diffuse and bright compact components.

@@

For the diffuse component, we use an all-sky radio foreground model from \citet{deo08} to estimate the emission at 185~MHz. Since this map is predominantly based on the 408~MHz all-sky map of \citet{has82} which has an angular resolution of 0\fdg 85, we smoothed the 185~MHz diffuse emission map to the same resolution. However, to avoid any artifacts from sampling this map, we sample it at $\approx 27$\arcmin~intervals. Using the same source model, we also obtain the diffuse emission maps at 140~MHz and 160~MHz to estimate the spectral index map of the diffuse emission model.

>>>>

We characterize signatures of foreground radio emission and instrument configuration on the observed power spectra of redshifted H{\sc i} 21~cm line emission from the epoch of reionization (EoR) using a baseline (antenna pair) based ``delay spectrum'' analysis technique. We use the 128-tile Murchison Widefield Array (MWA) instrument configuration for our studies. % Our analysis is centered around 185~MHz with 30.72~MHz bandwidth at 80~kHz frequency resolution. 
We simulate the delay spectrum response of interferometers, for the first time, to an all-sky foreground model consisting of diffuse (on scales $\gtrsim\,$0\fdg 85) and compact Galactic and extragalactic emission. By tying model parameters to those of a subset of first season of observing with the MWA, the ratio of our predictions to observations, subject to expected errors, are consistent with unity at $\gtrsim$~90\% confidence level. A wedge-shaped region of foreground contamination and a spillover from this contamination into relatively foreground--free regions ({\it EoR window}) in the delay spectrum due to spectral properties of foreground emission and the instrument is confirmed. Due to shortening of antenna spacings caused by projection effects, diffuse emission produces edge-heavy features in the wedge even on large antenna spacings, hitherto not predicted. Features in the inner regions of the wedge are predominantly due to compact foreground objects. This results in a ``pitchfork'' shaped delay spectrum. We also provide a practical diagnostic tool for data analysis that helps in identifying interferometers with high foreground contamination.

@@

After the CMB epoch, the period in the history of the universe often referred to as the {\it dark ages}, followed the epoch of reionization (EoR). This was a period of non-linear growth of density perturbations and astrophysical evolution. Studying the epoch of reionization holds the key to understanding this evolution. Recently, observing the redshifted 21~cm spin transition of neutral hydrogen has emerged as a very promising experiment to fill the gaps in our understanding of the universe's history.

@@

In this paper, in one of the most detailed characterizations to date, using an interferometer (antenna pair) based delay spectrum technique, we demonstrate the signatures of unsubtracted diffuse and compact foreground models in delay spectrum, a quantity closely related to the sought power spectrum. The simplicity of this technique and usage of basic interferometric visibility data fourier transformed along frequency (delay spectrum) on individual interferometers offer very effective diagnostics for scheduling observations and data analysis, and an insightful view onto foreground signatures in the EoR H{\sc i} power spectrum. We explore the diversity in these signatures as a function of observation and instrument parameters and confirm our findings with data obtained from the MWA. This simple technique allows us to predict foreground signatures, hitherto not predicted.

We have organized this paper as follows. \S\ref{sec:delay-spectrum} defines the delay spectrum and its relation to the power spectrum. \S\ref{sec:instrument} and \S\ref{sec:obsparms} describe the MWA instrumental parameters and observation parameters used in this study respectively. Analysis of data obtained with the MWA is described in \S\ref{sec:data-analysis}. In \S\ref{sec:modeling}, we describe the building blocks of our delay spectrum modeling including the models used for the antenna power pattern and foreground. Here, we also establish that the results of modeling and data analysis agree with each other. In \S\ref{sec:delay-spectrum-analysis}, we present a detailed analysis of the signatures of foregrounds on the modeled delay spectrum and its dependence on instrumental and observation parameters. In \S\ref{sec:fg-grading}, we use our understanding to grade interferometers based on the amount of foreground contamination estimated by the delay spectrum technique. In \S\ref{sec:summary}, we present our summary and conclusions.

@@

Interferometer array data known as {\it visibilities}, $V_f(\overline{\mathbf{u}},f)$, represent correlations between time--series of electric fields measured by different antenna pairs with separation vectors $\overline{\mathbf{x}}$ and then Fourier transformed along time axis to obtain a spectrum along frequency ($f$) axis. Here, $\overline{\mathbf{u}}\equiv \overline{\mathbf{x}}f/c \equiv \overline{\mathbf{x}}/\lambda$, $c$ is the speed of light and $\lambda$ is the wavelength. $\overline{\mathbf{u}}$ is related to transverse spatial frequency modes ($\overline{\mathbf{k}}_\perp$) of brightness distribution of the sky as $\overline{\mathbf{k}}_\perp \equiv 2\pi\overline{\mathbf{u}}/D(z)$, where $D(z)$ is the transverse comoving distance at a redshift $z$. If $I(\overline{\mathbf{l}},f)$ is the emission at different frequencies on the sky as a function of direction unit vector ($\overline{\mathbf{l}}$) specified as direction cosines, $V_f(\overline{\mathbf{u}},f)$ represents the Fourier decomposition of $I(\overline{\mathbf{l}},f)$ attenuated by the antenna's angular power pattern $A(\overline{\mathbf{l}},f)$, into transverse spatial frequency modes. Taking into account the instrumental bandpass weights $W_f(f)$, this relation can be written as:
\begin{align}\label{eqn:obsvis}
  V_f(\overline{\mathbf{u}},f) &= \int A(\overline{\mathbf{l}},f)\,I(\overline{\mathbf{l}},f)\,W_f(f)\,e^{-i2\pi \overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}\,\dif^2 \overline{\mathbf{l}}.
\end{align}

We define the {\it delay spectrum} $V_\eta(\overline{\mathbf{u}},\eta)$ to be the inverse Fourier transform of $V_f(\overline{\mathbf{u}},f)$ along the frequency coordinate:
\begin{align}\label{eqn:delay-transform}
  V_\eta(\overline{\mathbf{u}},\eta) &= \int V_f(\overline{\mathbf{u}},f)\,W'_f(f)\,e^{i2\pi\eta f}\,\dif f,
\end{align}
where, $W'_f(f)$ is a frequency window weighting function which can be chosen to control the quality of the delay spectrum \citep{thy13,ved12}. $\eta$ represents the signal delay between the antenna pairs in an interferometer given by:
\begin{equation}\label{eqn:delay}
  \eta = \frac{\overline{\mathbf{x}}\cdot\overline{\mathbf{l}}}{c} = \frac{\overline{\mathbf{u}}\cdot\overline{\mathbf{l}}}{f}.
\end{equation}
The terms delay and lag are used interchangeably in this paper.

Since we are studying a redshifted H{\sc i} spectral line, $f$ is a measure of cosmological distance along the line of sight. Consequently, $\eta$ is a measure of line of sight spatial frequency mode, $k_\parallel$. Thus, $|V_\eta(\overline{\mathbf{u}},\eta)|^2$ is directly related to the spatial power spectrum of redshifted H{\sc i} distribution, $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$. However, foreground objects emit in radio frequencies and contaminate the signals from the redshifted H{\sc i}. Due to their spatial and spectral properties, $V_\eta(\overline{\mathbf{u}},\eta)$ also contains contamination from foreground radio sources \citep{thy13,tro12,mor12,bow09}. These contaminations have to be characterized precisely in order to reduce their impact on redshifted H{\sc i} power spectrum detection sensitivity.

In summary, $V_\eta(\overline{\mathbf{u}},\eta)$, is closely related to visibilities, $V_f(\overline{\mathbf{u}},f)$, the basic data blocks measured by an interferometer, as seen from equation~\ref{eqn:delay-transform}. It captures all the effects of EoR H{\sc i} signal corruption caused by foregrounds and the instrument. At the same time, it is very closely related to the sought power $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$ in $k$--modes. For our analysis, we use an equivalent quantity, $V_\eta(\overline{\mathbf{x}},\eta) \equiv V_\eta(\lambda\overline{\mathbf{u}},\eta)$. For purposes of illustration, we often use a simplified quantity $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$.

We now describe in detail an important ingredient, the MWA instrument.

@@

We use the 128-tile layout of the MWA \citep{bea12} and choose a center frequency of 185~MHz with a bandwidth of 30.72~MHz divided into 384 frequency channels each of width 80~kHz. The choice of center frequency translates to $z\approx 6.68$ for the 21~cm spin flip transition of H{\sc i}. The bandwidth was chosen to be close to the maximum instantaneous value of the MWA in order to maximize the resolution in delay coordinates and obtain detailed characteristics of delay spectra from foregrounds. We restrict our analysis to the innermost antenna spacings with $|\overline{\mathbf{x}}| \leq x_\textrm{max} = 200$~m, or $|\overline{\mathbf{u}}| \leq u_\textrm{max} = 123$ at a frequency of 185~MHz. These interferometers are responsive to transverse spatial structures on scales $\gtrsim 1/u_\textrm{max} \approx 28$\arcmin. Hereafter, we use the terms tile and antenna interchangeably to refer to the MWA tile.

The bandpass weights in the frequency window of the MWA are obtained through a 8-tap Polyphase Filter Bank (PFB) using a Kaiser window with the $\beta$ parameter set to 5. The bandpass weights consist of coarse frequency channels of width 1.28~MHz each of which contains 16 finer frequency channels of width 80~kHz. Our analysis, therefore, contains 24 coarse channels. As the data are subsequently bandpass calibrated, the bandpass weights are normalized to unity. But the edges of the coarse channels are noisy and aliased from adjacent coarse channels. Thus, we remove one fine channel from each side of the coarse channel edges from our data. This constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}. The bandpass properties used in this paper are illustrated in figure~\ref{fig:bandpass}.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/bandpass_properties}
\caption{Bandpass properties used in modeling and analysis of delay spectrum of MWA visibilities. The filled circles joined by dotted lines show the 8-tap PFB bandpass shapes obtained using a Kaiser window with parameter $\beta=5$. The {\it plus} symbols joined by dashed lines are the bandpass correction factors. The solid line shows the flagged (gaps) and gain--corrected bandpass shape, $W_f(f)$. These shapes are repeated every 1.28~MHz for the entire bandwidth of 30.72~MHz centered around 185~MHz. \label{fig:bandpass}}
\end{figure}

Thermal noise in simulated visibilities is estimated assuming a system temperature of $T_\textrm{sys}=90$~K per polarization for all frequency channels for all interferometers throughout the course of observation. We take into account the increase in thermal uncertainty in frequency channels that results due to aforementioned bandpass correction.

@@

One of the primary targets for EoR observing with the MWA is a patch of sky centered around RA = 00h 00m 00.00s, Dec = -30\arcdeg$\,$00\arcmin$\,$00\farcs 0. This is one of the regions of low foreground emission within 27\arcdeg$\,$ from the zenith. The MWA tracks a patch of sky through antenna beams formed and steered electronically by controlling delay settings of an array of dipoles in a MWA tile. These beamformer delay settings can be changed only in discrete steps. Accordingly, observations were made in a ``drift and shift'' (D-S) mode. Figure~\ref{fig:pointings} shows the direction of pointing of MWA tiles during the course of the observation. The sky is allowed to drift for a certain period of time (usually $\sim 30$~mins) before the discrete shifts in the beamformer delay settings position the beam to be centered again on the same patch of sky. This process is repeated throughout the course of the observation $\approx 4.86$~hours. The LST of two snapshots we chose for our study are shown as vertical lines at -1.92~hours (wrapped by 360\arcdeg from 22.08~hours) and 0.09~hours which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings respectively. These pointings (phased array power patterns) are centered at RA = 4\fdg 387, Dec = -29\fdg 86 and RA = -28\fdg 8, Dec = -26\fdg 701, whereas the visibilities themselves are phased to zenith corresponding to RA = -28\fdg 8, Dec = -26\fdg 701, and RA = 1\fdg 35, Dec = -26\fdg 701 respectively.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/custom_pointings.eps}
\caption{MWA tile beam pointing directions during the course of the observation. The x-axis refers to the Local Sidereal Time (LST) in hours. The axis on the left refers to longitudes, namely, Right Ascension (RA) and Hour Angle (HA) in degrees. Negative values of RA, HA, and LST are to be interpreted as having been wrapped around by 360\arcdeg or 24 hours. The axis on the right refers to the declination of the pointing direction. The RA, HA, and declination are plotted with solid, dashed and dotted line styles respectively. A ``drift and shift'' scheme is used. The sky is allowed to drift for $\sim 30$~mins before the beamformer delay settings change in a discrete step to center the beam around RA = 00h 00m 00.00s, Dec = -30\arcdeg$\,$00\arcmin$\,$00\farcs 0. Dot--dashed vertical lines show the LST of off-zenith and zenith pointings used in our study. \label{fig:pointings}}
\end{figure}

Throughout this paper, $\overline{\mathbf{u}}$ and $\overline{\mathbf{x}}$ are assumed to be on a coordinate system aligned with the local east and north (along local meridian) directions at the MWA site. For the conventions of Fourier transform and its inverse we use in this paper, in accordance with equation~\ref{eqn:delay}, signals from the sky towards east and north are recorded with positive delays on eastward and northward oriented antenna spacings respectively. For instance, at the beginning of the observation, the Galactic center is in the westward sky just about to set. Signals from this direction are recorded with a negative delay on interferometers whose antenna spacings are oriented towards the east. Similarly, signals from directions eastward of the local meridian arrive with positive delays on eastward oriented antenna spacings.

For geometrical intuition, we restrict the orientation ($\theta_b$, measured anti-clockwise from East) of all antenna spacings to lie in the range $-67\fdg 5 \leq \theta_b < 112\fdg 5$. Interferometers with antenna spacings oriented in the other half-plane measure conjugate visibilities with delays of equal magnitude but of opposite sign and hence are ignored in our analysis.

@@

\subsection{Imaging and deconvolution}\label{sec:FHD}

{\bf Describe the FHD data analysis up to the point of getting calibrated visibilities here.}

@@

We obtain the delay spectrum of these calibrated visibilities using equation~\ref{eqn:delay-transform} while choosing $W'_f(f)$ to be a {\it Blackman--Harris} window function. Due to periodic gaps in frequency bandpass occurring in intervals of 1.28~MHz (see figure~\ref{fig:bandpass}), the delay spectrum is expected to contain harmonics of actual foreground emission repeated at intervals of 0.78~$\mu$s. Hence, we have employed deconvolution by {\it CLEAN} algorithm \citep{tay99} along the delay axis \citep{par09,par12} to rid the delay spectra of such artifacts. The convolving kernel for the algorithm is given by the inverse Fourier transform of the instrumental bandpass shape.

\subsection{Delay Spectrum after Delay Deconvolution}\label{sec:data-delay-spectrum}

We show results of delay spectrum from MWA data after deconvolution along delay axis in figure~\ref{fig:fhd_data}. Boundaries of the {\it foreground wedge} are also shown. The {\it off--zenith} pointing has emission notably higher than in the {\it zenith} pointing inside the wedge boundaries. This is shown later to be due to response of the interferometer array to the bright Galactic center and Galactic plane in the westward sky. Consequently, the contamination into the {\it EoR window} is also higher. Emission in {\it zenith} pointing is more centrally concentrated due to phasing of power pattern to the zenith, whereas the {\it off--zenith} pointing has a power pattern phased eastward of zenith and is also responsive to emission from angles far away from zenith.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/multi_baseline_fhd_delay_spectrum_snapshots.eps}
\caption{Delay spectrum amplitude, $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$ (in units of Jy~Hz), obtained with MWA data for {\it off--zenith} (top) and {\it zenith} (bottom) pointings. White lines mark the boundaries of {\it foreground wedge} determined by the horizon limit and antenna spacing. The features in the wedge appear to be brighter and the spillover contamination appears to be higher in the {\it off--zenith} pointing relative to the {\it zenith} pointing due to the presence of brighter foregrounds such as the Galactic center and the Galactic plane. The logarithmic color scale (shown at the right) is common to both panels.\label{fig:fhd_data}}
\end{figure}

Despite delay deconvolution, we still see leakage beyond the cleaned regions indicating the inability of the deconvolution algorithm to fit the data perfectly. This is predominant at short baselines. This is because the maximum delay envelope (boundary of the wedge) consists of mixed emission from large portions of the sky into a narrow range of delays and the deconvolution algorithm does not have sufficient support in delay space to fit the delay spectrum. However, we note that the purpose of deconvolution is to rid the delay spectra of instrumental artifacts to the best extent possible in order to see the underlying signatures of foreground emission. It is not our intention in this paper to use the deconvolution as a foreground removal tool. Hence, our discussion and conclusions still hold despite imperfections in deconvolution.

@@

We describe power pattern and foreground models we have used in modeling the measured delay spectra.

\subsection{Power pattern}\label{sec:power_pattern}

The power pattern of a MWA tile, $A(\overline{\mathbf{l}},f)$ in equation~\ref{eqn:obsvis}, over the entire hemisphere is analytically modeled based on a 4$\times$4 phased array of isotropic radiators at a height of 0.3~m above an infinite ground plane. We also assume that the delay corrections during the phased addition of dipole voltages in the beamformer suffer from random fluctuations of rms 0.05~ns. The models of power pattern used in data analysis take into account effects of mutual coupling of dipoles in the tile besides finite ground plane effects. Our models were found to be in good agreement with those used in the data analysis.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_powerpattern_0.3m_ground_custom.eps}
\caption{Models of MWA tile power pattern in Right Ascension ($\alpha$) and Declination ($\delta$) coordinates for the off--zenith (top) and zenith pointings (bottom) at 185~MHz. An MWA tile is modeled as a 4$\times$4 array of isotropic radiators placed 0.3~m above the ground plane. Random fluctutations of rms 0.05~ns have been added to delay corrections during the phased addition of voltages from these isotropic radiators. The logarithmic color scale (shown at the right) is common to both panels. \label{fig:power_pattern}}
\end{figure}

@@

An instrument such as the MWA has a very wide field of view ($\gtrsim 20$\arcdeg at 185~MHz) for imaging purposes. In the context of EoR H{\sc i} power spectrum, it has been known that unsubtracted forground sources anywhere in the visible hemisphere up to the horizon directly contaminate the spatial frequency modes in the power spectrum. In addition, this contamination also spills over into the relatively cleaner regions, called {\it EoR window}, due to spectral properties of the instrument and the foregrounds \citep{thy13,pob13,ved12,par12}. Thus, it is important to consider an all-sky model for foreground objects in evaluating the features seen in the power spectrum instead of restricting to the primary field of view.

\citet{bea13} estimated the sensitivity of the MWA to EoR H{\sc i} power spectrum detection taking into account theremal noise effects in the {\it Eor window}. \citet{thy13} estimated the sensitivity taking into account the spillover from foreground contamination from unsubtracted extragalactic point sources in the {\t EoR window} besides thermal noise. In this paper, we include the diffuse Galactic emission for enhancing our understanding of foreground signatures in the power spectrum. We use an all-sky foreground emission model that consists of diffuse and bright compact components.

@@

For the diffuse component, we use an all-sky radio foreground model \citep{deo08} to estimate the emission at 185~MHz. Since this map is predominantly based on the 408~MHz all-sky map of \citet{has82} which has an angular resolution of 0\fdg 85, we smoothed the 185~MHz diffuse emission map to the same resolution. However, to avoid any artifacts from sampling this map, we sample it at $\approx 27$\arcmin~intervals. Using the same source model, we also obtain the diffuse emission maps at 170~MHz and 200~MHz to estimate the spectral index map of the diffuse emission model.

A low resolution version of the diffuse foreground model used is show in figure~\ref{fig:DSM}. Contours of the MWA tile power pattern shown in figure~\ref{fig:power_pattern} are also overlaid. Of notable significance in the {\it off--zenith} pointing is the presence of a portion of the Galactic plane and the bright Galactic center in the visible hemisphere in the westward sky and the MWA tile's power receptivity is significantly high ($\gtrsim 0.125$) in that direction. In the {\it zenith} pointing, the Galactic plane has almost set and the contour level of power pattern in that direction is at least 16 times lesser.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/dsm.eps}
\caption{Sky brightness temperature (in K) of the diffuse foreground model at 185~MHz in Right Ascension ($\alpha$) and Declination ($\delta$) coordinates visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels. Contours of power pattern shown in figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.001953125, 0.0078125, 0.03125, 0.125, and 0.5. The Galactic center and a portion of the Galactic plane are prominently visible during the {\it off--zenith} pointing and the MWA tile power receptivity is significant ($\gtrsim 0.125$) in that direction. In contrast, emission from the Galactic plane in {\it zenith} pointing is significantly lesser. \label{fig:DSM}}
\end{figure}
